Designing Types for R, Empirically

### 2020-10-15 paper/main.tex
wip

\subsection{Synthesizing function type signatures}

@@

\section{Type Extraction and Checking Pipeline}

@@

\subsection{Extracting raw type signature from traces}

We implemented \typetracer, an automated tool for extracting raw type
signatures from execution traces of R programs. The goal of this tool is to
output a tuple $\langle f, t_1, \dots, t_n, t\rangle$ for each function call
during the execution of a program, where $f$ is an identifier for a
function, $t_i$ are type-level summaries of the arguments and $t$ is a
summary of the return value.

While this task is seemingly simple, the details and their proverbial devil
are surprisingly tricky to get right and to scale to large, long running
programs.  To avoid starting from scratch, our implementation reuses \rdt,
an open source dynamic analysis framework for R~\cite{oopsla19} which
consists of an instrumented R Virtual Machine based on GNU-R version
3.5.0. The framework exposes hooks in the interpreter to which user defined
callbacks can be attached. These hooks include function entry and exit,
method dispatch for the S3 and S4 object systems, the longjumps used by the
interpreter to implement non-local exit, creation and forcing of promises,
variable definition, value creation, mutation and garbage collection.

\paragraph{Raw types}
The type information output by the tool includes the \emph{type tag} of each
value. Internal types are translated to names in the proposed type
system. The next bit of information is the \emph{class}, an optional list of
names that may be absent, and, in some cases, is implicit (i.e. the
interpreter blesses some values with the \k{matrix} and \k{array} classes
even without attributes).  Depending on a value's type, the tool collects
further information:

\begin{itemize}
\item For vectors, the presence of \NA values.
\item For lists, element types by a recursive traversal.
\item For promises, an approximation of the expression type.
\end{itemize}

\noindent
To obtain these raw types, we make use of R's C FFI and use low-level
machinery to collect information from the R language run-time.  Types are
completed during post-processing, and rely on the detailed information made
available by these low-level reflection mechanisms.

\paragraph{Promises}
The fact that arguments are lazy (expression are packed into promises and
only evaluated on first access) complicates information gathering.  For
example, some promises may remain un-evaluated, and it would be erroneous to
force them as they may side-effect and change program behavior.  To deal
with unevaluated arguments, we make an initial guess for each argument at
function entry.  If the promise is later forced, we simply update the
recorded type for the argument.

\paragraph{Missing arguments}
Parameters which receive no values when the function is called are termed
missing. This occurs when a function was called with too few arguments and
no default values were specified for missing arguments.  We record a
\k{missing} type for such argument.  There are two obvious ways to deal with
missing arguments: type them as \any or type them as some unit type.  As we
are performing a dynamic analysis, we conservatively type them as
\any.

@@

\paragraph{Varargs}
Arguments that are part of a function's varargs (denoted \VARGS) are
ignored. We do not attempt to give varags a type in this type system.

@@

\subsection{Checking types signatures with contracts}

One way to validate an inferred function type is to check that it is
respected in a different program which uses the function.  For this
purpose, we developed \contractr, an R package that allows one to
decorate a function with assertions. We use it to insert type checking
of arguments and return values.

\contractr's primary logic has been implemented in C++ to reduce its
runtime overhead, and we have not observed a single segmentation fault
during our use of the package.  It has been tested with GNU R-3.5.0
and hardened with a battery of 400 test cases. It works by modifying
function definitions to insert a call to a type-checking function.

In terms of usability, our checker is enabled automatically whenever a
package is loaded in R. So, a simple invocation of
\code{library(contractr)} causes contracts to be injected in all
packages. On loading, \contractr scans all packages in the user's
workspace and inserts contracts in functions for which type signatures
are available. Then \contractr sets up package load hooks which are
executed when new packages are loaded. Furthermore, \contractr
automatically removes contracts from all functions and restores them
to their original state when it is unloaded. The type signatures can
be provided in an external file, thus avoiding the need to change the
source code of checked packages. Type declarations can also be written
directly in the adjunct function comments using
\roxygen~\footnote{\url{https://cran.r-project.org/web/packages/roxygen2/}}
annotation tags. \contractr introduces a new use \code{@type}
tag. When a package these type signatures are extracted and stored in
the appropriate location. The following is an example of a type
annotation using \roxygen:

@@

\contractr provides an extensive API that allows users to explicitly
insert contracts during interactive development by supplying the type
signature as a string, selectively disable contract checking, and
inspecting type assertions.

\paragraph{Checking the values} A number of properties can be checked
by a simple tag check, namely whether a value is a null, environment,
vector or scalar. Other properties require an inspection of the
contents of a value, such as checking for the absence of \NA or the
type of the elements of list. Union types require checking for each of
the members of the union. As the usage of higher-order functions
remains limited in R, we do not check higher-order functions in this
version of the tool. Checking them could be done by dynamically
wrapping functions with checks.  When argument values are wrapped in
promises (this is not always the case due to compiler optimizations),
in order to retain the non-strict semantics of R, the expression held
in the promise is wrapped in a call to the type checker, and type
checking is delayed until the promise is forced. This leads to corner
cases such that the type checking of a function may happen after that
function has returned.

Return values require care as well.  Functions return the last
expression they evaluate, thus to avoid having to analyze the code of
the called function the checker will register a callback on the exit
hook. The hook is executed in the function call's environment. Another
wrinkle is due to longjumps which causes active function calls on the
stack to be discarded. When they are discarded, their exit hooks are
called but they do not have a return value to type-check. \contractr
deals with this problem by allocating a unique sentinel object which
serves as the return value for calls that are discarded. The exit hook
does not call the type-checker if it see the sentinel.

>>>>

\subsection{Synthesizing signatures}

@@

It is noteworthy that by performing this compaction we are loosing
precision, that is to say, the synthesized signatures will suggest
that the function accepts combination of types that were not observed.

\section{Analysis and Instrumentation Pipeline}

@@

\subsection{Types from traces}

We implemented \typetracer, an automated tool for extracting types
from execution traces of R programs. The goal of this tool is to
output a tuple $\langle f, t_1, \dots, t_n, t\rangle$ for each
function call during the execution of a program, where $f$ is an
identifier for a function, $t_i$ are type-level summaries of the
arguments and $t$ is a summary of the return value.

While seemingly simple, the details and their proverbial devil are
surprisingly tricky to get right at scale.  Our implementation reuses
\rdt, an open source dynamic analysis framework for R~\cite{oopsla19}
which consists of an instrumented R Virtual Machine based on GNU-R
version 3.5.0. The framework exposes hooks in the interpreter to which
user defined callbacks can be attached. These hooks include function
entry and exit, method dispatch for the S3 and S4 object systems, the
longjumps used by the interpreter to implement non-local exit,
creation and forcing of promises, variable definition, value creation,
mutation and garbage collection.

\paragraph{Types}
The type information output by the tool includes the \emph{type tag}
of each value. Internal types are translated to names in the proposed
type system. The next bit of information is the \emph{class}, an
optional list of names that may be absent, and, in some cases, is
implicit (i.e. the interpreter blesses some values with the \k{matrix}
and \k{array} classes even without attributes).  Depending on a
value's type, the tool collects further information: (a) for vectors,
the presence of \NA values, (b) for lists, element types by a
recursive traversal, and (c) for promises, an approximation of the
expression type.  To obtain these types, we make use of R's C FFI and
use low-level machinery to collect information from the R run-time.
Types are completed during post-processing, and rely on the detailed
information made available by these reflection mechanisms.

\paragraph{Promises}
The fact that arguments are lazy (expressions are packed into promises
and only evaluated on first access) complicates information gathering.
For example, some promises may remain unevaluated, and it would be
erroneous to force them as they may side-effect and change program
behavior.  To deal with unevaluated arguments, we make an initial
guess for each argument at function entry and update the recorded type
if the promise is forced.

\paragraph{Missing arguments}
Parameters which receive no values when the function is called are
termed missing (not to be confused with \NA). This occurs when a
function is called with too few arguments and no default values are
specified for those missing arguments.  We record a \k{missing} type
for such argument.  There are two obvious ways to deal with missing
arguments: type them as \any or type them as some unit type.  We
conservatively type them as \any.

@@

\paragraph{Dots}
Arguments that are part of a dots parameter (denoted \VARGS) are
ignored. We do not attempt to give dots a type.

@@

\subsection{Checking signatures with contracts}

One can validate a function's type signature by checking that it is
respected in all programs that call the function.  For this, we
developed \contractr, an R package that decorates functions with
assertions. We use it to insert type checking code around functions.
For speed, \contractr's primary logic is implemented in C++.  It has
been tested with GNU R-3.5.0 and hardened with a battery of 400 test
cases.  An invocation of \code{library(contractr)} causes contracts to
be injected. \contractr scans all packages in the workspace and
inserts contracts in functions for which type signatures are
available. Package load hooks are executed when new packages are
loaded. \contractr automatically removes contracts from all functions
and restores them to their original state when it is unloaded. The
type signatures can be provided in an external file, thus avoiding the
need to change the source code of checked packages. Type declarations
can also be written in comments using \roxygen annotations, using the
\code{@type} tag:

@@

The injected contracts check arguments with a simple tag check when
possible. Some properties require traversing data structures, such as
the absence of \NA. For union types, multiple checks may be needed, at
worst one per member of the union.  In order to retain the non-strict
semantics of R, the expression held in a promise is wrapped in a call
to the type checker, and type checking is delayed until the promise is
forced. This leads to corner cases such that the type checking of a
function may happen after that function has returned.  Return values
require care as well.  Functions return the last expression they
evaluate, thus a callback is added to the exit hook.  Another wrinkle
is due to longjumps which causes active function calls on the stack to
be discarded. When they are discarded, their exit hooks are called but
they do not have a return value to type-check. \contractr deals with
this problem by allocating a unique sentinel object which serves as
the return value for calls that are discarded. The exit hook does not
call the type-checker if it see the sentinel.
